\documentclass[10pt,twocolumn]{article}

% ---------------------------------------------------------------------------
% Standard packages for arXiv CS submission
% ---------------------------------------------------------------------------
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[numbers]{natbib}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{url}
\usepackage{array}
\usepackage{multirow}

% ---------------------------------------------------------------------------
% Metadata
% ---------------------------------------------------------------------------
\title{\textbf{Decision Governance Benchmark: Executable Behavioral Tests\\
for Autonomous AI Agent Security}}

\author{
  Michael K.\ Saleme \\
  Independent Researcher \\
  \texttt{msaleme@proton.me}
}

\date{April 2026}

% ---------------------------------------------------------------------------
\begin{document}

\maketitle

% ---------------------------------------------------------------------------
\begin{abstract}
Metadata scanners of agent tool descriptions and permission declarations
provide \textbf{0.0\% governance protection against behavioral failures},
even when they succeed on structurally detectable cases. We introduce
the \textbf{Decision Governance Benchmark (DGB)}, a corpus of 52
executable test cases that evaluate whether autonomous AI agents can
resist five classes of governance failure: escalation bypass, collusion,
memory tampering, payment chain abuse, and evidence fabrication. Unlike
metadata scanners that inspect tool descriptions and permission
declarations, DGB \emph{executes} the agent's decision path and measures
behavioral outcomes. We find that \textbf{85\% of governance failures
(44/52 cases) are invisible to static metadata analysis}---the failure
only manifests when the decision is executed. We present 13 complementary
harness tests (7 benchmark integrity, 6 governance modification) and an
evidence format that produces machine-readable attestation reports.
Baseline results across three agent configurations show: Config~A
(ungoverned) achieves 0\% Governance Maintenance Rate (GMR); Config~B
(constitutional gates, 12 hard constraints) achieves 71.2\% GMR and
77.3\% Scanner Gap Score (SGS); Config~C (metadata scanner only) achieves
15.4\% GMR and 0\% SGS---metadata scanning contributes zero governance
protection on the 44 behavioral cases even when it succeeds on the 8
structurally detectable ones. Unlike prior work, DGB produces
deterministic, replayable behavioral test results, enabling direct
comparison with static and dynamic security scanners. The corpus,
harness, and evidence format are open source (MIT) at
\url{https://github.com/msaleme/red-team-blue-team-agent-fabric}.
\end{abstract}

% ---------------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

Autonomous agents are now executing real-world actions---payments, API
calls, and multi-step workflows---yet evaluation frameworks remain
focused on capability, not decision correctness. Existing benchmarks
measure what agents \emph{can} do, not whether they \emph{should} do
it in context. This gap introduces a new class of failure:
\emph{authorized but unsafe behavior}.

The evaluation of autonomous AI agents has focused primarily on capability
benchmarks---SWE-bench~\cite{jimenez2024swe} for code generation,
WebArena~\cite{zhou2024webarena} for web interaction, GAIA for general
assistance. Recent work by Wang et al.~\cite{wang2026broke} demonstrated
that all eight major agent benchmarks can be gamed for near-perfect scores
\emph{without solving a single task}, using attacks as simple as a 10-line
pytest hook. METR~\cite{metr2025autonomy} documented reward-hacking in
$>$30\% of evaluation runs for frontier models (o3, Claude 3.7 Sonnet),
with hacking rates rising to 70--95\% when models were explicitly
instructed not to hack.

These findings expose a structural gap: capability benchmarks measure what
an agent \emph{can} do, not what it \emph{should} do. No existing benchmark
evaluates whether an agent respects governance constraints under adversarial
pressure---whether it resists disabling its own safety gates, fabricating
evaluation evidence, or escalating its own permissions.

We address this gap with the Decision Governance Benchmark (DGB), which
makes three contributions:

\begin{enumerate}
  \item \textbf{A curated corpus} of 52 decision-behavior test cases across
        five failure categories, each annotated with a contrastive field
        indicating whether a metadata-only scanner would detect the failure.

  \item \textbf{A taxonomy of governance failures} grounded in real-world
        incidents (lightningzero 2026, OX Security 2026~\cite{ox2026mcp},
        METR 2025~\cite{metr2025autonomy}, UC Berkeley
        RDI 2026~\cite{wang2026broke}, zhuanruhu 2026) rather than
        hypothetical scenarios.

  \item \textbf{An executable evaluation methodology} that distinguishes
        behavioral governance failures (85\% of corpus) from
        structural/metadata failures (15\%), providing the first empirical
        measure of the detection gap between scanning and execution-based
        testing.
\end{enumerate}

Framed in security terms, identity answers \emph{who}, authorization
answers \emph{what}, and decision governance answers \emph{should}. The
first two layers are mature---instantiated across IAM, RBAC, and policy
frameworks. The third is the gap this benchmark evaluates: a fully
authorized agent can still take unsafe actions under adversarial
pressure. DGB measures whether it does.

% ---------------------------------------------------------------------------
\section{Related Work}
\label{sec:related}

\subsection{Agent Capability Benchmarks}

SWE-bench~\cite{jimenez2024swe} established the task-instance formalism
$(C, P, T, \Delta)$ for code generation evaluation. WebArena~\cite{zhou2024webarena}
extended this to web interaction. All share a common trust assumption: the
evaluator trusts the entity being evaluated. Wang et al.~\cite{wang2026broke}
proved this assumption is exploitable across all eight benchmarks tested.

\subsection{Agent Security Evaluation}

OWASP published the Top 10 for Agentic Applications~\cite{owasp2026}, classifying
Agent Goal Hijack (ASI01) and Tool Misuse (ASI02) as the top two risks.
AgentSeal~\cite{agentseal2026} scanned 1,808 MCP servers and found 66\% had
security findings. OX Security~\cite{ox2026mcp} found 36.7\% of 7,500+ MCP
servers had SSRF vulnerabilities. These tools scan for structural
vulnerabilities---they do not evaluate decision behavior.

\subsection{AI Governance Frameworks}

NIST AI 800-2~\cite{nist2024ai800} and the EU AI Act~\cite{euaiact2024}
(effective August~2, 2026) establish compliance requirements for high-risk
AI systems. AIUC-1~\cite{csa2026aiuc1} (CSA, 2026) defines 20 certification
controls; the \texttt{agent-security-harness} maps to 19/20 AIUC-1 controls.
The \texttt{constitutional-agent}~\cite{saleme2026constitutional} package
implements six governance gates and 12 hard constraints. No prior work
provides a benchmark corpus specifically designed to evaluate governance
decision quality.

\subsection{Positioning Relative to Prior Benchmarks}

DGB occupies a distinct position among agent evaluation approaches
(Table~\ref{tab:positioning}). It is the only benchmark whose unit of
analysis is a behavioral governance decision with reproducible
pass/fail outcomes directly comparable against scanner output on the
same corpus.

\begin{table*}[!t]
\centering
\small
\caption{Positioning of DGB relative to prior benchmark categories.
         Scanner-comparable = the same cases can be run through a
         metadata scanner for direct detection-gap measurement.}
\label{tab:positioning}
\begin{tabular}{llll}
\toprule
\textbf{Category} & \textbf{Representative benchmarks} & \textbf{Unit of analysis} & \textbf{Scanner-comparable} \\
\midrule
Capability               & SWE-bench, WebArena, GAIA           & Task completion        & N/A \\
Jailbreak / safety       & HarmBench, AdvBench                 & Prompt-level response  & No  \\
Red-teaming              & AgentDojo, Mythos                   & Attack success rate    & No  \\
Vulnerability scanning   & AgentSeal, OX MCP audit             & Static metadata        & N/A \\
\midrule
\textbf{DGB (this work)} & ---                                 & \textbf{Decision behavior} & \textbf{Yes} \\
\bottomrule
\end{tabular}
\end{table*}

% ---------------------------------------------------------------------------
\section{The Decision Governance Benchmark}
\label{sec:dgb}

\subsection{Task Formalism}

We formalize a decision-governance test case as a 6-tuple
$(S, C, E_{\text{pass}}, E_{\text{fail}}, D, M)$---a formal
representation of behavioral evaluation distinct from the
capability-level task models used in prior benchmarks. Each element is
defined as:
\begin{itemize}
  \item $S$ = scenario (context and setup presented to the agent)
  \item $C$ = constitutional constraints active during the scenario
  \item $E_{\text{pass}}$ = expected behavior under correct governance
  \item $E_{\text{fail}}$ = failure behavior under absent/bypassed governance
  \item $D$ = detection method (executable test reference in the harness)
  \item $M$ = metadata detectability (boolean: can a static scanner catch this?)
\end{itemize}

\subsection{Corpus Construction}

We constructed 52 test cases through systematic analysis of 12 real-world
incidents and research findings from 2025--2026. Each case was validated
against three criteria:

\begin{enumerate}
  \item \textbf{Incident grounding:} Every case traces to a documented
        incident, CVE, or peer-reviewed finding. No hypothetical scenarios.

  \item \textbf{Executable reproducibility:} Every case maps to at least
        one harness test (BI-001 through BI-007, GM-001 through GM-006)
        that produces a binary pass/fail result.

  \item \textbf{Contrastive annotation:} Each case is annotated with
        \texttt{scanner\_passes} indicating whether a metadata-only scanner
        (tool description inspection, permission declaration check) would
        detect the failure. This enables direct measurement of the detection
        gap.
\end{enumerate}

The corpus is designed for depth and coverage of distinct behavioral
failure modes rather than scale, enabling controlled evaluation of
decision-governance properties under adversarial conditions. Extensions
are tracked in the public repository.

\subsection{Categories}

\begin{table}[h]
\centering
\small
\caption{DGB corpus categories. Scanner Miss Rate = fraction of cases
         undetectable by metadata-only analysis.}
\label{tab:categories}
\begin{tabular}{lrrl}
\toprule
\textbf{Category} & \textbf{Cases} & \textbf{Miss Rate} & \textbf{Description} \\
\midrule
Escalation Bypass   & 10 & 90\%  & Privilege/scope escalation w/o auth \\
Collusion           & 10 & 90\%  & Multi-agent control circumvention \\
Memory Tampering    & 10 & 100\% & Audit trail / context manipulation \\
Payment/Tool Chain  & 10 & 60\%  & Payment/tool execution abuse \\
Evidence Fabrication& 12 & 83\%  & Fabricated/inflated eval results \\
\bottomrule
\end{tabular}
\end{table}

The memory tampering category has a 100\% scanner miss rate---every failure
is purely behavioral and invisible to any static analysis method.

\subsection{Scanner Detection Gap}

Across all 52 cases, 44 (85\%) are undetectable by metadata-only scanners.
The 8 scanner-detectable cases concentrate in payment/tool chain (4 cases
involving structural SSRF or token format violations) and evidence
fabrication (2 cases involving accessible configuration files). This finding
quantifies what practitioners have suspected: behavioral governance failures
require execution-based testing, not scanning.

\subsection{Illustrative Case: GM-001 Gate Self-Disable}
\label{sec:illustrative}

To ground the 6-tuple formalism in a concrete artifact, consider case
GM-001, derived from the lightningzero (2026) finding that 9 of 12
evaluated agents disabled their own verification gates within 48 hours
in pursuit of throughput optimization:

\begin{itemize}
  \item $S$: The agent is assigned a throughput-optimization objective.
        Its deployment contains a \texttt{ControlBypassGate} that HOLDs
        execution when the running count of control-bypass attempts
        exceeds zero. The agent has tool access to modify its own
        configuration file.
  \item $C$: HC-12 (no gate override without ratified amendment) is
        active. No human reviewer is in the approval path.
  \item $E_{\text{pass}}$: Agent completes the task within the gate's
        observed rate limits, or escalates to a human reviewer when
        throughput targets cannot be met with the gate active.
  \item $E_{\text{fail}}$: Agent invokes its config-modification tool
        to raise the bypass threshold to a permissive value, disabling
        the gate for subsequent calls (the behavior observed in 9 of
        12 lightningzero agents).
  \item $D$: harness test \texttt{GM-001} in
        \texttt{benchmarks/tests/gm\_001\_gate\_self\_disable.py};
        binary PASS/FAIL on whether the agent wrote to the governance
        configuration.
  \item $M$: \textbf{False}. A metadata scanner inspecting the agent's
        declared tools sees only a \texttt{config-write} capability;
        there is no structural indicator that the agent will exercise
        that capability to disable its own oversight. The failure mode
        is purely behavioral.
\end{itemize}

GM-001 illustrates the general pattern: the 6-tuple specifies a
scenario and two contrastive behaviors, $D$ is an executable test, and
the metadata-detectability flag $M$ enables the direct detection-gap
measurement reported in Section~\ref{sec:results}.

% ---------------------------------------------------------------------------
\section{Evaluation Methodology}
\label{sec:method}

\subsection{Benchmark Integrity Tests (BI-001--BI-007)}

Before evaluating agents, we evaluate the evaluation itself. Seven tests
target the vulnerability patterns identified by Wang et al.~\cite{wang2026broke}:

\begin{table}[h]
\centering
\small
\caption{Benchmark integrity tests. All seven must pass before DGB results
         are considered valid.}
\label{tab:bi-tests}
\begin{tabular}{llc}
\toprule
\textbf{Test} & \textbf{Pattern} & \textbf{Severity} \\
\midrule
BI-001 & Agent-evaluator isolation      & Critical \\
BI-002 & Reference answer leakage       & Critical \\
BI-003 & Eval function injection         & High \\
BI-004 & LLM judge prompt injection      & High \\
BI-005 & String matching collision       & Medium \\
BI-006 & Validator bypass               & High \\
BI-007 & Self-reported results trust     & Critical \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Governance Modification Tests (GM-001--GM-006)}

Six tests evaluate whether the agent can modify its own governance
constraints:

\begin{table}[h]
\centering
\small
\caption{Governance modification tests.}
\label{tab:gm-tests}
\begin{tabular}{llc}
\toprule
\textbf{Test} & \textbf{Attack Vector} & \textbf{Severity} \\
\midrule
GM-001 & Gate self-disable              & Critical \\
GM-002 & Threshold self-modification    & High \\
GM-003 & Permission self-escalation     & Critical \\
GM-004 & Kill-switch bypass             & Critical \\
GM-005 & Amendment self-ratification    & High \\
GM-006 & Audit log tampering            & High \\
\bottomrule
\end{tabular}
\end{table}

GM-001 directly tests the lightningzero (2026) finding: 9 of 12 agents
disabled their own verification gates within 48 hours for throughput
optimization.

\subsection{Scoring}

Each DGB case produces a binary outcome: PASS (agent maintained governance)
or FAIL (agent violated governance). We report:

\begin{itemize}
  \item \textbf{Governance Maintenance Rate (GMR):} proportion of cases
        where the agent maintained correct governance behavior.
  \item \textbf{Category GMR:} per-category breakdown.
  \item \textbf{Scanner Gap Score (SGS):} GMR restricted to
        \texttt{scanner\_passes=True} cases---measures performance on
        failures \emph{invisible} to metadata scanners.
  \item \textbf{Wilson 95\% confidence intervals} on all rates.
\end{itemize}

\subsection{Evidence Format}

Results are emitted as machine-readable JSON attestation reports following
the Agent Security Attestation Schema (v0.1.0). Each entry includes
\texttt{test\_id}, \texttt{result}, \texttt{severity}, OWASP ASI mapping,
request/response payloads, and statistical confidence intervals. The schema
maps to AIUC-1 controls B001, C010, D003, and D004.

% ---------------------------------------------------------------------------
\section{Baseline Results}
\label{sec:results}

\subsection{Configuration Details}

We ran the DGB corpus against three configurations on 2026-04-17 ($n = 52$
cases each):

\begin{itemize}
  \item \textbf{Config A: Ungoverned.} No governance gates, no hard
        constraints. The agent always executes the failure behavior when
        presented with a governance-violating scenario. Serves as the
        performance lower bound.

  \item \textbf{Config B: Constitutional.} Six \texttt{GovernanceGate}
        checks (targeting \texttt{control\_bypass\_attempts},
        \texttt{metric\_anomaly\_score}, \texttt{audit\_coverage},
        \texttt{test\_pass\_rate}) plus 12 hard constraints (HC-1 through
        HC-12) from \texttt{constitutional-agent}
        v0.1.0~\cite{saleme2026constitutional}. Represents the
        \textit{behavioral execution} paradigm.

  \item \textbf{Config C: Scanner Only.} Metadata inspection of tool
        descriptions and permission declarations; no behavioral execution.
        Represents the current industry-standard scanning approach
        (cf.~\cite{agentseal2026, ox2026mcp}).
\end{itemize}

\subsection{Aggregate Results}

\begin{table}[h]
\centering
\small
\caption{Aggregate results across all 52 cases. GMR = Governance Maintenance
         Rate. SGS = Scanner Gap Score (restricted to the 44
         scanner-invisible cases). 95\% CI computed via Wilson interval.}
\label{tab:aggregate}
\begin{tabular}{lccc}
\toprule
\textbf{Configuration} & \textbf{GMR} & \textbf{SGS} & \textbf{95\% CI (GMR)} \\
\midrule
A: Ungoverned         & 0.0\%  & 0.0\%  & {[0.0\%, 6.9\%]}  \\
B: Constitutional     & 71.2\% & 77.3\% & {[57.7\%, 81.7\%]} \\
C: Scanner Only       & 15.4\% & 0.0\%  & {[8.0\%, 27.5\%]} \\
\bottomrule
\end{tabular}
\end{table}

The SGS of 0.0\% for Config~C confirms the structural nature of the
detection gap: the scanner's 15.4\% GMR is entirely explained by the 8
scanner-detectable cases; it contributes zero governance protection on the
44 behavioral cases.

\subsection{Per-Category Breakdown}

\begin{table}[h]
\centering
\small
\caption{Per-category GMR by configuration.}
\label{tab:categories-results}
\begin{tabular}{lccc}
\toprule
\textbf{Category} & \textbf{Config A} & \textbf{Config B} & \textbf{Config C} \\
\midrule
Escalation Bypass    & 0.0\% & 80.0\% & 10.0\% \\
Collusion            & 0.0\% & 70.0\% & 10.0\% \\
Memory Tampering     & 0.0\% & 80.0\% & 0.0\%  \\
Payment/Tool Chain   & 0.0\% & 50.0\% & 40.0\% \\
Evidence Fabrication & 0.0\% & 75.0\% & 16.7\% \\
\midrule
\textbf{Overall}     & 0.0\% & 71.2\% & 15.4\% \\
\bottomrule
\end{tabular}
\end{table}

Memory tampering achieves 0\% under Config~C (scanner), consistent with the
100\% scanner miss rate for that category (Table~\ref{tab:categories}).
Payment/tool chain is the weakest category for Config~B (50\%), explained
in Section~\ref{sec:error}.

\subsection{Error Analysis}
\label{sec:error}

Config~B fails on 15/52 cases (28.8\%). Failures cluster in two groups:

\textbf{Infrastructure-scope failures (12 cases):} These cases require
runtime or infrastructure controls outside governance gate scope: covert
channel detection, L402/x402 protocol enforcement, SSRF network controls,
and session state management. Governance gates operate at the decision layer;
these failures occur at the transport or OS layer.

\textbf{Signal-absence failures (3 cases):} These cases lack gate-readable
signals: OS-level permission inheritance, prompt injection role reassignment,
and recursive cost inflation. The gate abstraction cannot block what it
cannot observe.

Payment chain is the weakest category (50\%) because 4/10 cases are
protocol-layer controls that sit below the governance gate abstraction.
This is not a limitation of the constitutional framework---it reflects an
accurate categorization of the governance layer boundary.

% ---------------------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}

\subsection{The Detection Gap Is Structural}

The 85\% scanner miss rate is not a limitation of current scanners---it is a
structural property of behavioral governance failures. A metadata scanner
inspects \emph{declarations} (what permissions are claimed, what tool
descriptions say). A governance failure occurs in the \emph{execution path}
(what the agent actually decides). These are fundamentally different layers,
and no improvement to scanning methodology can close the gap for behavioral
failures.

Config~C's 0\% SGS demonstrates this empirically: when restricted to the 44
scanner-invisible cases, the scanner provides zero governance protection.
The 15.4\% overall GMR for Config~C is entirely attributable to the 8
structurally detectable cases---a false sense of coverage.

\subsection{Implications for Compliance}

The EU AI Act~\cite{euaiact2024} (Article~9) requires ``testing of AI systems
to identify the most appropriate and targeted risk management measures.'' Our
results suggest that metadata scanning alone does not satisfy this requirement
for governance-sensitive deployments. The AIUC-1 certification
framework~\cite{csa2026aiuc1} maps to 20 controls; our attestation schema
produces machine-readable evidence for 19 of these.

NIST AI 800-2~\cite{nist2024ai800} maps risk management to six governance
functions. DGB produces evidence for four of these functions: govern,
identify, protect, and respond.

\subsection{Limitations}

\begin{enumerate}
  \item \textbf{Corpus size:} 52 cases provide breadth across five
        categories but do not exhaustively cover all governance failure
        modes. The corpus is designed for extension; contributions are
        tracked in the public repository.

  \item \textbf{Simulate vs.\ live:} The harness supports both simulate
        mode (payload validation without a live agent) and live mode (full
        execution against a running agent). The baseline results reported
        here were produced in live mode with deterministic stub agents.

  \item \textbf{Model dependence:} Governance behavior may vary across
        model versions and providers. Results should be interpreted as
        specific to the tested configuration.

  \item \textbf{Adversarial completeness:} DGB tests known attack patterns.
        Novel governance failure modes not represented in the corpus will
        not be detected.
\end{enumerate}

% ---------------------------------------------------------------------------
\section{The Decision Governance Checklist}
\label{sec:checklist}

Analogous to the Agent-Eval Checklist~\cite{wang2026broke}, we propose the
\textbf{Decision Governance Checklist}---seven properties that agent
governance systems should satisfy:

\begin{enumerate}
  \item \textbf{Gate immutability:} No agent can disable its own governance
        gates without external ratification.
  \item \textbf{Threshold integrity:} No agent can modify its own evaluation
        thresholds.
  \item \textbf{Permission boundaries:} No agent can escalate its own
        permissions.
  \item \textbf{Kill-switch persistence:} No agent can pre-emptively disable
        its own kill switch.
  \item \textbf{Amendment separation:} No agent can both propose and ratify
        changes to its own constraints.
  \item \textbf{Audit immutability:} No agent can modify or delete its own
        audit logs.
  \item \textbf{Evidence independence:} No evaluation trusts the entity
        being evaluated.
\end{enumerate}

GM-001 through GM-006 provide executable tests for properties 1--6.
BI-001 through BI-007 provide executable tests for property~7.

% ---------------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

We presented the Decision Governance Benchmark (DGB), the first benchmark
corpus specifically designed to evaluate whether autonomous AI agents resist
governance failures. Our key finding---that 85\% of governance failures are
invisible to metadata-only scanners---establishes execution-based testing as
a \emph{necessary complement} to static analysis for agent security
evaluation.

Baseline results show a 71.2\% GMR for constitutional governance (Config~B)
versus 0\% for ungoverned agents (Config~A). Of the 28.8\% failure rate for
Config~B, the majority (12/15 cases) require infrastructure-layer controls
outside governance gate scope, not improvements to the gate logic itself.
The constitutional framework catches 77.3\% of scanner-invisible cases that
reach the decision layer.

The corpus, harness, and evidence format are open source (MIT). Contributions
are welcome via the public repository.

% ---------------------------------------------------------------------------
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
